---
title: "QC"
author: "Izaskun Mallona, Mark D. Robinson lab, UZH"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    code_download: true
    number_sections: true
    df_print: kable
    theme: lumen
params:
    multimodal_path: ""
    run_mode: ""
---



```{r}
suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(Seurat)
    library(scuttle)
    library(scater)
    library(ggplot2)
    library(rmarkdown)
})
```

This report processes:

```{r}
print(params$multimodal_path)
print(params$run_mode)
```

```{r}

ac <- function(col, alpha=1){
  apply(sapply(col, col2rgb)/255, 2, 
                     function(x) 
                         rgb(x[1], x[2], x[3], alpha=alpha))
}
```


```{r}
knitr::opts_chunk$set(fig.width = 5,
                      fig.height = 5,
                      error = TRUE,
                      echo = TRUE,
                      cache = TRUE,
                      include = TRUE,
                      dev = "png",
                      dev.args = list(type = "cairo-png"),
                      cache.lazy = FALSE,
                      warning = TRUE,
                      message = TRUE)

## render on error hook
knitr::knit_hooks$set(error = function(x, options) {
    ## if (exists(x)) {
    ##     cat(str(x))
    ## }

    knitr::knit_exit()
})
  
```

```{r}

## dp <- "/home/imallona/ebrunner_spectral/mixing_mouse_genome_only/multimodal_nidhi/"
dp <- params$multimodal_path
sces <- lapply(list.files(dp,
                          ".*sce.rds", recursive = TRUE),
               function(x) readRDS(file.path(dp, x)))

## combine sces
## sce <-  Reduce(cbind, sces)
for (i in 1:length(sces)) {
    colnames(sces[[i]]) <- paste0(dirname(list.files(dp,
                                                     ".*sce.rds", recursive = TRUE)),
                                          '_',
                                          colnames(sces[[i]]))
    colData(sces[[i]])$experiment <- mainExpName(sces[[i]])
    colnames(rowData(sces[[i]])) <- c("name", "type", "value")
    rownames(sces[[i]]) <- paste0(rowData(sces[[i]])$name, "__", rownames(sces[[i]]))
}

sce <-  Reduce(cbind, sces)
rm(sces)

```

# WTA

## Per cell and per feature QC {.tabset .tabset-fade .tabset-pills}

```{r, cache = FALSE}
sce <- addPerCellQC(sce,
                    assay.type = 'wta',
                    subsets = list(mito = grep('^mt-', rownames(rowData(sce)),
                                               value = TRUE, ignore.case = TRUE),
                                   capture = rowData(sce)$captured == 'captured'))

libsize.drop <- isOutlier(sce$total, nmads = 2, type = "both", log = TRUE, batch = sce$experiment)
feature.drop <- isOutlier(sce$detected, nmads = 2, type = "both", log = TRUE, batch = sce$experiment)
mito.drop <- isOutlier(sce$subsets_mito_percent, nmads = 2, type = "higher", log = FALSE,
                       batch = sce$experiment)
```

```{r, fig.width = 5, fig.height = 4, results = 'asis'}

cat('### ', 'Sum vs detected (by name)', ' \n\n')
plotColData(sce, x = "sum", y="detected", colour_by = "experiment")
cat('\n\n')

cat('### ', 'Highest expressed', ' \n\n')
plotHighestExprs(sce, exprs_values = "wta")
cat('\n\n')
```

## QC filtering

```{r}
sce <- sce[,!(libsize.drop | feature.drop | mito.drop )]
data.frame(ByLibSize = sum(libsize.drop),
           ByFeature = sum(feature.drop), 
           ByMito = sum(mito.drop),
           Remaining = ncol(sce))
```


## QC after filtering {.tabset .tabset-fade .tabset-pills}

### Per cell and per feature QC {.tabset .tabset-fade .tabset-pills}

```{r qualmetricsafter, fig.width = 5, fig.height = 4, results = 'asis'}

cat('#### ', 'Sum vs detected (by name)', ' \n\n')
plotColData(sce, x = "sum", y="detected", colour_by = c("experiment"))
cat('\n\n')

## cat('### ', 'Sum vs detected (by batch)', ' \n\n')
## plotColData(sce, x = "sum", y="detected", colour_by = c("batch"))
## cat('\n\n')

## cat('### ', 'Sum vs detected (by color)', ' \n\n')
## plotColData(sce, x = "sum", y="detected", colour_by = c("color"))
## cat('\n\n')


cat('### ', 'Highest expressed', ' \n\n')
plotHighestExprs(sce, exprs_values = "tso_off_and_ontarget_unique")
cat('\n\n')
```


### Overall quality metrics after QC (hists) {.tabset .tabset-fade .tabset-pills}


```{r hists_plots_after, results = 'asis'}

for (id in unique(colData(sce)$experiment)) {
    par(mfrow=c(2,2), mar=c(5.1, 4.1, 0.1, 0.1))

    curr <- subset(sce, , experiment == id)
    
    cat('#### ', id, ' \n\n') 
    hist(curr$total/1e3, xlab="Library sizes (thousands)", main="", 
         breaks=20, col="grey80", ylab="Number of cells")
    abline(v = attr(libsize.drop, 'thresholds')['lower', id]/1e3, col = 'blue')
    abline(v = attr(libsize.drop, 'thresholds')['higher', id]/1e3, col = 'blue')

    hist(curr$detected, xlab="Number of expressed genes", main="", 
         breaks=20, col="grey80", ylab="Number of cells")
    abline(v = attr(feature.drop, 'thresholds')['lower', id], col = 'blue')
    abline(v = attr(feature.drop, 'thresholds')['higher', id], col = 'blue')

    ## hist(curr$altexps_spikes_percent, xlab="ERCC proportion (%)",
    ##      ylab="Number of cells", breaks=20, main="", col="grey80")
    ## abline(v = attr(spike.drop, 'thresholds')['higher', id], col = 'blue')

    hist(curr$subsets_mito_percent, xlab="Mitochondrial proportion (%)", 
         ylab="Number of cells", breaks=20, main="", col="grey80")
    abline(v = attr(mito.drop, 'thresholds')['higher', id], col = 'blue')
    cat('\n\n')
}

```

### Overall quality metrics after QC (scatter) {.tabset .tabset-fade .tabset-pills}

```{r library_sizes_after, , results = 'asis'}
for (id in unique(colData(sce)$experiment)) {
    cat('#### ', id, ' \n\n') 
    par(mfrow=c(2,2), mar=c(5.1, 4.1, 0.1, 0.1))

    curr <- subset(sce, , experiment == id)
    
    plot(density(curr$total/1e3),
         xlab = "library sizes (thousands)",
         main = '')

    rug(curr$total/1e3)

    plot(y = curr$total/1e3,
         x = curr$subsets_mito_percent,
         ## col = ac(curr$color, 0.5),
         pch = 20,
         ylab = 'library size (thousands)',
         xlab = 'mitochondrial proportion (%)',
         ## col = ac('black', 0.5)
         )

    plot(y = curr$total/1e3,
         x = curr$detected,
         ## col = ac(curr$color, 0.5),
         pch = 20,
         ylab = 'library size (thousands)',
         xlab = 'number of genes')
    ## col = ac('black', 0.5)
    ## )

    ## plot(y = curr$total/1e3,
    ##      x = curr$altexps_spikes_percent,
    ##      col = ac(as.numeric(as.factor(curr$name)), 0.5),
    ##      pch = 20,
    ##      ylab = 'library size (thousands)',
    ##      xlab = "ERCC proportion (%)",
    ##      ## col = ac('black', 0.5)
    ##      )

    plot(y = curr$subsets_mito_percent,
         x = curr$detected,
         pch = 20,
         xlab = 'number of genes',
         ylab = 'mitochondrial proportion (%)',
         col = ac('black', 0.5))

    cat('\n\n')
} 
```

## Quick clustering (WTA) {.tabset .tabset-pills}

```{r}
meta <- as.data.frame(t(assay(altExp(sce, 'tso_ontarget_multi'))))[rownames(colData(sce)),]
colnames(meta) <- paste0('captured_', colnames(meta))

so <- CreateSeuratObject(counts = assay(sce, 'wta'),
                         project = "wta",
                         meta.data = as.data.frame(cbind(colData(sce), meta)),
                         min.cells = 0, min.features = 0,
                         assay = 'RNA')

rm(meta)
```

```{r}
so <- NormalizeData(object = so, normalization.method = 'LogNormalize', scale.factor = 10000)
so <- FindVariableFeatures(so, selection.method = "vst", nfeatures = 2000)

so <-  ScaleData(so, features = rownames(so))

so <- RunPCA(object = so,
             features = VariableFeatures(so),
             verbose = FALSE,
             npcs = 50)

```


### Loadings

```{r, fig.width = 12, fig.height = 8}
VizDimLoadings(object = so, dims = 1:4)
```

### Elbow

```{r, fig.width =4, fig.height = 4}
print(ElbowPlot(object = so, ndims = 50))

```

```{r}
dims <- 1:30
```

```{r, message = FALSE}

so <- FindNeighbors(object = so, dims = dims)

## (resolutions <- seq(0.2, 1.4, by = 0.6))
resolutions <- c(0.1, 0.5)

for (res in resolutions) {
    so <- FindClusters(object = so, resolution = res)
}

```

### PCA {.tabset .tabset-fade .tabset-pills}

```{r, fig.width = 7, fig.height = 4, results='asis'}

## print(colnames(so@meta.data))

layers <- c("experiment",
            grep('snn_res', colnames(so@meta.data), value = TRUE))

for (item in layers){

    cat('#### ', item, ' \n\n')
    print(DimPlot(object = so, reduction = "pca", group.by = item, shuffle = TRUE) +
          ggtitle(item) +
          theme(legend.position="right", aspect.ratio = 1))
    cat('\n\n')
}

cat('#### RNA counts \n\n')
FeaturePlot(object = so, reduction = "pca", features = 'nCount_RNA') +
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')

cat('#### Number of features \n\n')
FeaturePlot(object = so, reduction = "pca", features = 'nFeature_RNA')+
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')

cat('#### Percent mitochondrial reads \n\n')
FeaturePlot(object = so, reduction = "pca", features = 'subsets_mito_percent')+
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')


captures <- grep('captured', colnames(so@meta.data), value = TRUE)
for (item in captures){
     tryCatch({   
        cat('#### ', item, ' \n\n')
        print(FeaturePlot(object = so, reduction = "pca", features = item)+
              theme(legend.position="right", aspect.ratio = 1))
        cat('\n\n')
}, error = function(x) print(x))
}


```

### UMAPs {.tabset .tabset-fade .tabset-pills}

```{r, fig.width = 8, fig.height = 4, message = FALSE}
so <- RunUMAP(object = so, reduction = "pca", 
              dims = dims)
```


```{r, fig.width = 7, fig.height = 4, results='asis'}

for (item in layers){
    cat('#### ', item, ' \n\n')
    print(DimPlot(object = so, reduction = "umap", group.by = item, shuffle = TRUE) +
          ggtitle(item) +
          theme(legend.position="right", aspect.ratio = 1))
    cat('\n\n')
}

cat('#### RNA counts \n\n')
FeaturePlot(object = so, reduction = "umap", features = 'nCount_RNA') +
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')

cat('#### Number of features \n\n')
FeaturePlot(object = so, reduction = "umap", features = 'nFeature_RNA') +
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')

cat('#### Percent mitochondrial reads \n\n')
FeaturePlot(object = so, reduction = "umap", features = 'subsets_mito_percent') +
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')

for (item in captures){
tryCatch({
    
    cat('#### ', item, ' \n\n')
    print(FeaturePlot(object = so, reduction = "umap", features = item)+
        theme(legend.position="right", aspect.ratio = 1))
    cat('\n\n')
    }, error = function(x) print(x))
}

```

# TSO (on- and offtarget, unique) {.tabset .tabset-pills}

## Per cell and per feature QC {.tabset .tabset-pills}

```{r}
## colData(sce) <-  colData(sce)[,'experiment']

## most of the columns are WTA-based QC anyway
colnames(colData(sce)) <- paste0('wta_', colnames(colData(sce)))
colData(sce)$wta_experiment <- colData(sce)$experiment

sce <- addPerCellQC(sce,
                    assay.type = 'tso_off_and_ontarget_unique',
                    subsets = list(mito = grep('^mt-', rownames(rowData(sce)),
                                               value = TRUE, ignore.case = TRUE),
                                   capture = rowData(sce)$captured == 'captured'))

```


```{r, fig.width = 5, fig.height = 4, results = 'asis'}

cat('### ', 'Sum vs detected (by name)', ' \n\n')
plotColData(sce, x = "sum", y="detected", colour_by = "experiment")
cat('\n\n')

cat('### ', 'Highest expressed', ' \n\n')
plotHighestExprs(sce, exprs_values = "tso_off_and_ontarget_unique")
cat('\n\n')
```


## QC after WTA filtering {.tabset .tabset-fade .tabset-pills}

### Per cell and per feature QC {.tabset .tabset-fade .tabset-pills}

```{r, fig.width = 5, fig.height = 4, results = 'asis'}
cat('#### ', 'Sum vs detected (by name)', ' \n\n')
plotColData(sce, x = "sum", y="detected", colour_by = c("experiment"))
cat('\n\n')

cat('### ', 'Highest expressed', ' \n\n')
plotHighestExprs(sce, exprs_values = "tso_off_and_ontarget_unique")
cat('\n\n')
```


### Overall quality metrics after QC (hists) {.tabset .tabset-fade .tabset-pills}

```{r}
ac <- function(col, alpha=1){
  apply(sapply(col, col2rgb)/255, 2, 
                     function(x) 
                         rgb(x[1], x[2], x[3], alpha=alpha))
}
```

```{r, results = 'asis'}

for (id in unique(colData(sce)$experiment)) {
    par(mfrow=c(2,2), mar=c(5.1, 4.1, 0.1, 0.1))

    curr <- subset(sce, , experiment == id)
    
    cat('#### ', id, ' \n\n') 
    hist(curr$total/1e3, xlab="Library sizes (thousands)", main="", 
         breaks=20, col="grey80", ylab="Number of cells")

    hist(curr$detected, xlab="Number of expressed genes", main="", 
         breaks=20, col="grey80", ylab="Number of cells")

    ## hist(curr$altexps_spikes_percent, xlab="ERCC proportion (%)",
    ##      ylab="Number of cells", breaks=20, main="", col="grey80")
    ## abline(v = attr(spike.drop, 'thresholds')['higher', id], col = 'blue')

    hist(curr$subsets_mito_percent, xlab="Mitochondrial proportion (%)", 
         ylab="Number of cells", breaks=20, main="", col="grey80")

    cat('\n\n')
}

```

### Overall quality metrics after WTA QC (scatter) {.tabset .tabset-fade .tabset-pills}

```{r, results = 'asis'}
for (id in unique(colData(sce)$experiment)) {
    cat('#### ', id, ' \n\n') 
    par(mfrow=c(2,2), mar=c(5.1, 4.1, 0.1, 0.1))

    curr <- subset(sce, , experiment == id)
    
    plot(density(curr$total/1e3),
         xlab = "library sizes (thousands)",
         main = '')

    rug(curr$total/1e3)

    plot(y = curr$total/1e3,
         x = curr$subsets_mito_percent,
         ## col = ac(curr$color, 0.5),
         pch = 20,
         ylab = 'library size (thousands)',
         xlab = 'mitochondrial proportion (%)',
         ## col = ac('black', 0.5)
         )

    plot(y = curr$total/1e3,
         x = curr$detected,
         ## col = ac(curr$color, 0.5),
         pch = 20,
         ylab = 'library size (thousands)',
         xlab = 'number of genes')
    ## col = ac('black', 0.5)
    ## )

    ## plot(y = curr$total/1e3,
    ##      x = curr$altexps_spikes_percent,
    ##      col = ac(as.numeric(as.factor(curr$name)), 0.5),
    ##      pch = 20,
    ##      ylab = 'library size (thousands)',
    ##      xlab = "ERCC proportion (%)",
    ##      ## col = ac('black', 0.5)
    ##      )

    plot(y = curr$subsets_mito_percent,
         x = curr$detected,
         pch = 20,
         xlab = 'number of genes',
         ylab = 'mitochondrial proportion (%)',
         col = ac('black', 0.5))

    cat('\n\n')
} 
```


## Quick clustering (TSO on- and offtarget unique) {.tabset .tabset-pills}

```{r}
meta <- as.data.frame(t(assay(altExp(sce, 'tso_ontarget_multi'))))[rownames(colData(sce)),]
colnames(meta) <- paste0('captured_', colnames(meta))

so <- CreateSeuratObject(counts = assay(sce, 'tso_off_and_ontarget_unique'),
                         project = "tso unique",
                         meta.data = as.data.frame(cbind(colData(sce), meta)),
                         min.cells = 0, min.features = 0,
                         assay = 'RNA')

rm(meta)
```

```{r}
so <- NormalizeData(object = so, normalization.method = 'LogNormalize', scale.factor = 10000)
so <- FindVariableFeatures(so, selection.method = "vst", nfeatures = 2000)

so <-  ScaleData(so, features = rownames(so))

so <- RunPCA(object = so,
             features = VariableFeatures(so),
             verbose = FALSE,
             npcs = 50)

```


### Loadings

```{r, fig.width = 12, fig.height = 8}
VizDimLoadings(object = so, dims = 1:4)
```

### Elbow

```{r, fig.width =4, fig.height = 4}
print(ElbowPlot(object = so, ndims = 50))

```

```{r}
dims <- 1:30
```

```{r, message = FALSE}

so <- FindNeighbors(object = so, dims = dims)

## (resolutions <- seq(0.2, 1.4, by = 0.6))
resolutions <- c(0.1, 0.5)

for (res in resolutions) {
    so <- FindClusters(object = so, resolution = res)
}

```

### PCA {.tabset .tabset-fade .tabset-pills}

```{r, fig.width = 7, fig.height = 4, results='asis'}

## print(colnames(so@meta.data))

layers <- c("experiment",
            grep('snn_res', colnames(so@meta.data), value = TRUE))

for (item in layers){

    cat('#### ', item, ' \n\n')
    print(DimPlot(object = so, reduction = "pca", group.by = item, shuffle = TRUE) +
          ggtitle(item) +
          theme(legend.position="right", aspect.ratio = 1))
    cat('\n\n')
}

cat('#### RNA counts \n\n')
FeaturePlot(object = so, reduction = "pca", features = 'nCount_RNA') +
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')

cat('#### Number of features \n\n')
FeaturePlot(object = so, reduction = "pca", features = 'nFeature_RNA')+
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')

cat('#### Percent mitochondrial reads \n\n')
FeaturePlot(object = so, reduction = "pca", features = 'subsets_mito_percent')+
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')

captures <- grep('captured', colnames(so@meta.data), value = TRUE)
for (item in captures){
     tryCatch({   
        cat('#### ', item, ' \n\n')
        FeaturePlot(object = so, reduction = "pca", features = item)+
            theme(legend.position="right", aspect.ratio = 1)
        cat('\n\n')
}, error = function(x) print(x))
}


```

### UMAPs {.tabset .tabset-fade .tabset-pills}

```{r, fig.width = 8, fig.height = 4, message = FALSE}
so <- RunUMAP(object = so, reduction = "pca", 
              dims = dims)
```


```{r, fig.width = 7, fig.height = 4, results='asis'}

for (item in layers){
    cat('#### ', item, ' \n\n')
    print(DimPlot(object = so, reduction = "umap", group.by = item, shuffle = TRUE) +
          ggtitle(item) +
          theme(legend.position="right", aspect.ratio = 1))
    cat('\n\n')
}

cat('#### RNA counts \n\n')
FeaturePlot(object = so, reduction = "umap", features = 'nCount_RNA') +
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')

cat('#### Number of features \n\n')
FeaturePlot(object = so, reduction = "umap", features = 'nFeature_RNA') +
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')

cat('#### Percent mitochondrial reads \n\n')
FeaturePlot(object = so, reduction = "umap", features = 'subsets_mito_percent') +
    theme(legend.position="right", aspect.ratio = 1)
cat('\n\n')

for (item in captures){
tryCatch({
    
    cat('#### ', item, ' \n\n')
    FeaturePlot(object = so, reduction = "umap", features = item)+
        theme(legend.position="right", aspect.ratio = 1)
    cat('\n\n')
    }, error = function(x) print(x))
}

```


# TSO ontarget

## Raw counts {.tabset .tabset-pills}

```{r, results = 'asis'}

print(rownames(altExp(sce, 'tso_ontarget_multi')))

colData(altExp(sce, 'tso_ontarget_multi')) <- colData(sce)

for (experiment in unique(colData(altExp(sce, 'tso_ontarget_multi'))$experiment)) {
    cat('### ', experiment, ' \n\n')
    
    idx <- colData(altExp(sce, 'tso_ontarget_multi'))$experiment == experiment
    plot(colSums(assay(altExp(sce[,idx], 'tso_ontarget_multi'))),
         main = experiment,
         xlab = 'Cell (index)')

    cat('\n\n')
}
```

## Detection rates {.tabset .tabset.pills}

These are percentage of cells expressing each (captured) feature.

```{r, results = 'asis'}
for (experiment in unique(colData(altExp(sce, 'tso_ontarget_multi'))$experiment)) {
    cat('### ', experiment, ' \n\n')
    
    idx <- colData(altExp(sce, 'tso_ontarget_multi'))$experiment == experiment
    print(kable(apply(assay(altExp(sce[,idx], 'tso_ontarget_multi')),
          1,
          function(x) sum(x > 0)/sum(idx) * 100)))
         

    cat('\n\n')
}

```

# Session info

```{r}
sessionInfo()
```
